### 2014-09-11 paper/paper.Rmd
intro and section 1

Reducing inequities in education outcomes is of great interest to politicians and the public, alike. Dissecting the causes of inequality is a much tougher challenge with no panacea. Nonetheless, tackling inequality in educational outcomes is essential from both moral and economic viewpoints. The Programme for International Student Assessment (PISA) is a triennial survey conducted by the Organization for Economic Cooperation and Development (OECD) with a rotating additional emphasis on one of mathematics, reading, and science. In 2012, the additional emphasis was on mathematics. All 34 member countries of the OECD and 31 partnering countries and economies participated in the survey. This represents over 80% of the global economy. The OECD estimates that if all students could reach a level-2 proficiency in mathematics -- a level-2 proficiency means that a student can only handle the ``simplest and most obvious tasks" -- it would add US$200 trillion to the world's collective GDP.

The OECD describes the PISA survey as ``the world's global metric for quality, equity and efficiency in school education" (citep[p. 3]{PISA2014a}). The goal of the PISA survey is to assess the workforce readiness of 15-year old students. Nearly 500,000 students were tested across the 65 countries and economies. Students are examined on how well they can apply the knowledge they learned in school to applications outside of school. The scores reported range between 0-1000. Additional information about the students, parents, and schools is also collected. The students completed a questionnaire providing information about themselves, their homes, their schools, and a variety of psychological views regarding factors they believe affects their performance in school. Schools principals responded to a questionnaire covering the school system and learning experiences for students. In some countries, parents completed a questionnaire requesting information about their perceptions regarding the school system, expectations for their child, and their involvement in their child's schooling.

The data was made available by the OECD as part of a data challenge for the useR! 2014 conference. Entries to the competition can be found at http://beta.icm.edu.pl/PISAcontest/. We learned a lot from this data, and won one of the prizes. This article is a description of the seven main things we learned about 15-year olds, their math, science and reading skills, and some of the factors that affect these skills.

@@

For each country, average scores on math and reading are calculated separately for boys and girls. These averages are differenced to compute the "gap" between boys and girls. T-statistics were calculated to determine the statistical significance of the difference. The results are plotted in the graphs below. Each dot corresponds to a difference between girls and boys scores, for each country. Color indicates statistical significance (blue in favor of boys, pink in favor of girls and green means there was no significant difference). The size of the dot indicates the proportion of boys in the sample size. For most countries it is roughly equal, but some countries have an imblance, for example, Malaysia has more girls tested than boys. The maps color the countries by the presence of a gender gap.

The countries are sorted from biggest gap to smallest gap. Most countries have a gender gap in math scores in favor of boys. Colombia has the biggest gender gap, about 30 points on average boys do better than girls. It should be noted that this is small, though, because this 30 points is out of a possible 1000 points.  Four countries have gender gaps in favor of girls, and even more surprising, these are middle eastern, or Muslim countries.

Conversely, though, when we look at reading scores the story is entirely different. Universally, girls score significantly better than boys. Again, the size of the gap is small, between 10 and 80 points.

Although, not shown here, it also should be noted that this is the difference between the average scores for boys and girls. There is a slightly different story for each country if we look at the top math score for each geneder only, where we would notice that the top score in the USA is achieved by a girl, and is substantially higher than the highest boys' score. There is far more individual variability than the overall average differences.

@@

>>>>

Reducing inequities in educational outcomes is of great interest to politicians and the public, alike. Dissecting the causes of inequality is a much tougher challenge with no panacea. Nonetheless, tackling inequality in educational outcomes is essential from both moral and economic viewpoints. The Programme for International Student Assessment (PISA) is a triennial survey conducted by the Organization for Economic Cooperation and Development (OECD) with a rotating emphasis on one of mathematics, reading, or science. In 2012, the emphasis was on mathematics. All 34 member countries of the OECD and 31 partnering countries and economies participated in the survey. This represents over 80% of the global economy. The OECD estimates that if all students could reach a level-2 proficiency in mathematics -- a level-2 proficiency means that a student can only handle the ``simplest and most obvious tasks" -- it would add US$200 trillion to the world's collective GDP.

The OECD PISA Results in Focus report describes the survey as ``the world's global metric for quality, equity and efficiency in school education". The goal of the PISA survey is to assess the workforce readiness of 15-year old students. Nearly 500,000 students were tested across 65 countries and economies. Students were examined on how well they can apply the knowledge they learned in school to applications outside of school. The  reported scores range between 0-1000. Information about the students, parents, and schools is also collected. The students completed a questionnaire providing information about themselves, their homes, their schools, and a variety of psychological views regarding factors they believe affect their performance in school. School principals responded to a questionnaire covering their school system and learning experiences for their students. In some countries, parents completed a questionnaire requesting information about their perceptions regarding the school system, expectations for their child, and their involvement in their child's schooling.

The data was made available by the OECD as part of a data challenge for the useR! 2014 conference. Entries to the competition can be found at http://beta.icm.edu.pl/PISAcontest/. We learned a lot from this data, and won one of the prizes. This article is a description of the seven main things we learned about 15-year olds, their math, science, and reading skills, and some of the factors that affect these skills.

@@

For each country, average scores on math and reading are calculated separately for boys and girls. These averages are differenced to compute the "gap" between boys and girls. Unadjusted t-statistics were calculated to determine the statistical significance of the difference. The results are plotted in the graphs below. Each dot corresponds to a difference between girls' and boys' scores, for each country. Color indicates statistical significance (blue in favor of boys, pink in favor of girls, and green means there was no significant difference). The size of the dot indicates the proportion of boys in the sample size. For most countries it is roughly equal, but some countries have an imblance, for example, Malaysia has more girls tested than boys. The maps color the countries by the presence of a gender gap.

The countries are sorted from biggest gap to smallest gap. Most countries have a gender gap in math scores in favor of boys. Colombia has the biggest gender gap, about 30 points on average boys do better than girls. It should be noted that this is small, though, because this 30 points is out of a possible 1000 points.  Five countries have gender gaps in favor of girls, and even more surprising, these are middle eastern or Muslim countries.

Conversely, though, looking at reading scores, the story is entirely different. Universally, girls score significantly better than boys. Again, the size of the gap is small, between 10 and 80 points.

Although not shown here, it also should be noted that this is the difference between the average scores for boys and girls. There is a slightly different story for each country if we look at the top math score for each gender only. We would notice that the top score in the USA is achieved by a girl, and is substantially higher than the highest boys' score. There is far more individual variability than the overall average differences: who you are is more important than your demographic.

@@

### Reference

OECD (2014) "Pisa 2012 results in focus: What 15-year-olds know and what they can do with what they know." Technical Report, Organization for Economic Development and Cooperation.
